---
title: "Understanding the structure of rodent species assemblages and land use change on the occurrence of the rodent host of Lassa Fever."
author:
  - name: "David Simons"
    url: https://github.com/diddrog11/rodent_trapping
    affiliation: The Royal Veterinary College, London, UK
    affiliation_url: https://www.rvc.ac.uk
    orcid_id: 0000-0001-9655-1656
date: "`r Sys.Date()`"
output:
  # distill::distill_article:
  #   self_contained: false
  #   toc: true
  # html_document
  # word_document
  pdf_document
bibliography: citations.bib
categories:
  - Lassa
header-includes:
    - \usepackage{caption}
---

```{r setup, include=FALSE, quietly=TRUE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

library(flextable)
library(here)
library(tidyverse)

```

# Abstract

*Lassa mammarenavirus*, the causative agent of Lassa fever is endemic to Eastern Sierra Leone. The principal reservoir species (*Mastomys natalensis*), is considered abundant in human dominated habitats, however, rodent species' assemblages in this context are not well described. We conducted three monthly small-mammal trapping to describe these rodent assemblages, their structure and associations with land use. We model the effect of land use on rodent species occurrence along a land use gradient and produce species distribution maps of the study region to understand current and potential *Lassa mammarenavirus* spillover hazard.

We found that *M. natalensis* were more likely to occur in areas of human habitation and agricultural settings although there was important variability between different study villages. *M. natalensis* co-occurred with other other small-mammals including the two invasive rodent species (*Rattus rattus* and *Mus musculus*). We found evidence that the presence of *M. musculus*, bot not *R. rattus* reduced the probability of occurrence of *M. natalensis*, this finding potentially goes some way to explain the observation of lower than expected human cases of Lassa fever from urban settings in endemic regions. Other native rodent species' within these assemblages were found to diversify into distinct habitat niches. Species distribution maps identified areas of expected occurrence and non-occurrence of our species of interest *M. natalensis* and potential geographic isolation of populations.

We identify a complex system within rodent species assemblages co-located with human communities in Eastern Sierra Leone. We show the habitat occupancy patterns for each species of interest and use these observations to produce species distribution maps that explain the limited geographic radiation of outbreaks of Lassa fever. We anticipate that this data will help inform higher resolution models of rodent distributions across West Africa, which are of particular importance for rodent zoonotic diseases such as Lassa fever. These data highlight the spatially heterogeneous distribution of important rodent species with implications for public health interventions to reduce the impact of Lassa fever.

# Introduction

Lassa fever, caused by *Lassa mammarenavirus* (LASV) is an endemic zoonotic infectious disease in West Africa, with an estimated 100,000-900,000 annual human infections [@mccormick_prospective_1987; @basinski_bridging_2021]. The majority of these remain undetected, up-to 80% of these are pauci- or asymptomatic infections [@mccormick_prospective_1987]. However, outcomes in confirmed cases remains poor, with a reported case fatality rate of 18.5% among confirmed cases identified between 2017 and 2020 in Nigeria [@yaro_infection_2021]. Changing land-use and climate are hypothesised to increase the suitable area for both the primary reservoir of LASV (*Mastomys natalensis*) and environmental suitability for the virus itself, together increasing opportunities for viral spillover into growing human populations [@redding_environmental-mechanistic_2016; @redding_geographical_2021; @klitting_predicting_2021]. Lassa fever is currently considered endemic in eight West African countries (Nigeria, Guinea, Sierra Leone, Liberia, Mali, Benin, Ghana and Togo) [@world_health_organisation_lassa_2022]. The location, scale and frequency of Lassa fever outbreaks in human communities is highly spatially heterogeneous. In Sierra Leone, Lassa fever is typically reported from rural settings, however, the potential drivers of this have not been systematically investigated.

The persistence, or lack of persistence, of LASV infection in rodent communities is one factor driving spatial differences of outbreaks in human populations. In a well mixed, isolated rodent population  (i.e. rural settings), once introduced LASV would be expected to rapidly infect susceptible populations. Infected rodents do not develop clinical symptoms, with antibody mediated resistance developing following acute infection. This could lead to local elimination of the virus from these rodent populations. Conversely in highly-connected, diverse and poorly mixed rodent communities (i.e. urban settings) LASV persistence would be expected to be prolonged with viral transmission maintained through the rapid reproduction rate of rodents maintaining a large population of rodents susceptible to infection. In this way the suitability of both habitat type, climate and resource availability for both the primary reservoir and incidental reservoirs can drive the observed spatial clustering of Lassa fever outbreaks within Sierra Leone.

*M. natalensis* is found in 13 of 14 continental West African nations (not reported from Gambia) and all other sub-Saharan African nations [@iucn_iucn_2016]. It is considered a commensal rodent species and is abundant in and around areas of human-dominated landscapes where it is considered a pest species [@leirs_productivity_1993]. The introduction of non-native commensal rodent species (i.e. *Rattus rattus* and *Mus musculus*) has led to increased competition for resources and displacement of *M. natalensis* from locations within its natural range [@cuypers_no_2017; @garba_spatial_2014]. Population dynamics within this reservoir species, correlated with resource availability and rainfall, are associated with outbreaks of Lassa fever in human populations [@redding_geographical_2021]. Few studies to date have used longitudinal, high intensity rodent trapping to characterise rodent species assemblages in Lassa fever endemic regions (**ideally reference scoping review manuscript**). Understanding the true spatial distribution of *M. natalensis* and their population dynamics in the context of competing rodent species is vital to guide investigations of the epidemiology of Lassa fever [@basinski_bridging_2021]. Further, description of rodent abundance and diversity along land use gradients are required to better understand the spatio-temporal hazard of Lassa fever outbreaks under changing land use pressures [@klitting_predicting_2021]. Together, this information can be used to guide the implementation of contextually relevant public health responses, allocation of healthcare resources and the identification of suitable sites for future Lassa fever vaccine studies.

Here, we use data from a repeat, standardised, rodent trapping survey based in the Lassa fever endemic region of Eastern Sierra Leone and conducted along a land-use gradient to understand the association between land-use on the occurrence of *M. natalensis* and more generally, the rodent species assemblage structure. We first report the occurrence of rodent species at our trapping sites and describe these species assemblages, producing networks of co-located species. Second, we model the association of land-use with the probability of species occupancy at trapping sites. Third, we model the probability of co-occurrence between different species detected in our study to understand competition between small mammal species in our study region. Together these analyses further our understanding of rodent species assemblage structures in this heterogeneous landscape and the hazard of Lassa fever outbreaks based on host species occurrence.

# Methods

## Study area

We conducted rodent trapping at 7 trapping sites within 4 villages in the Lassa fever endemic zone of the Eastern Province of Sierra Leone. We surveyed the rodent community in forested, fallow, agricultural and areas of human occupation (within and outside of homes) along an anthropogenic land use gradient. Current land-use was classified at the level of trapping site as Agriculture, Fallow-land and Forested. Traps placed within villages were further defined as being within human dwellings (Village inside) and outside of permanent structures (Village outside). The village sites were enrolled based on accessibility to the sites during all seasons, discussions with the Lassa fever outreach team at Kenema Government Hospital and acceptability of the protocol to the village community. Villages and trapping sites were selected to be representative at the study level for land use in Eastern Sierra Leone.

```{r trap-locations, layout = "l-page", fig.cap = "Figure 1. A) Locations of study villages in Eastern Province, Sierra Leone. B) Trap sites in Lalehun village. C) Trap sites in Seilama village. D) Trap sites in Lambayama. E) Trap sites in Baiama village. The trap success rate in each trap sites for all study visits is shown through the colour associated with each trap site. Single points represent the location of a single trap."}

knitr::include_graphics(here("output", "grid_locations.pdf"))

```

Trap sites were geo-located for repeated trapping activities, changes to land-use at the trapping site were recorded at each visit. Within each study site 49 individual Sherman traps (**size and reference**) were baited with a locally produced mixture of oats, palm oil and dried fish for 4 consecutive nights. Each morning the traps were checked and closed for the day prior to re-baiting during the evening. Individual traps placed during each visit were associated with standardised 7m by 7m grid squares (49m^2^) within trapping sites. The number of trap nights and the number of trapped small-mammals by species within each grid square were aggregated.

### Rodent sampling

Trapped rodents were sedated with halothane and euthanised prior to obtaining morphological measurements and samples of blood and tissue (**reference to RVC and local ethics approval**) following published guidance [@fichet-calvet_chapter_2014]. Rodents were sexed based on external and internal genitalia. Age estimation was performed through description of the rodents reproductive status (identification of perforate or imperforate vagina, scarring from prior embryo development, current pregnancy status or descent of testes and seminal vesicle development) and weighing of dried eye lenses. Carcasses were disposed and processed in the field to eliminate risk of pathogen transmission.

Molecular identification to species was performed on dried blood spots that were stored at -20°C until processing. Genomic DNA was extracted using QIAGEN DNAeasy kits as per the manufacturers instructions [ref]. DNA extracts were amplified using platinum *Taq* polymerase (Invitrogen) and cytochrome B primers. DNA amplification was assessed through gel electrophoreisis with successful amplification products undergoing Sanger sequencing. Obtained sequences were compared using BLAST against NCBI records for rodent cytochrome B.

## Statistical analysis

### Rodent occurrence and species assemblage structure

We obtained 40,152 trap-nights over 10 trapping visits between 2020-11-30 and 2023-04-28. Trapping effort was assessed using species accumulation curves (Supplementary figure 1.), suggesting adequate effort to detect rodent species within each village site. We constructed detection/non-detection histories for all identified rodent species, assigning "1" when the species was detected and "0" otherwise, aggregated to standardised grid squares. We augmented data by creating all-zero detection histories of rodent species that have been previously described as occurring in the region and were never recorded in our study.

We describe species assemblages at multiple geographic scales. First, all species identified across all village sites and land-use types. Second, all species identified within a village. Third, all species identified within a single land-use type across multiple trapping sites and villages.

### Rodent species networks

To understand potential contact between individual small mammal species we produce a buffer zone, with a radius of 50m, around each trapped individual. All other individuals trapped within this area are identified as potentially co-located individuals. The `sf` package in the R statistical computing language was used for geospatial manipulation and analysis [@pebesma_simple_2018; @r_core_team_r_2021] The proportion of these individuals from each species are calculated and compared to a null model where co-located species are randomly allocated with equal probability. We then compare whether the proportion of each co-located is greater or lesser than this null model.

### Estimating the effect of land use on species occurrence and richness

To understand the association of small mammal species and land-use types we adopted a Bayesian multi-species occupancy framework to model occupancy from rodent trapping data in the presence of incomplete detection. Models were defined using the `spOccupancy` package in the R statistical computing language [@doser_spoccupancy_2022]. This approach models the true presence or absence (${z}$) of a species ($i$), at site ($j$) as arising from a Bernoulli process (Equation 1.). Where ${\psi}_{j}$ is the probability of occurrence of a species at a site during a given replicate. This is modelled using a logit link where $\beta_{i}$ are the regression coefficients, including an intercept that describes the effects of covariates $\boldsymbol{x}_{i,j}$ with ${}^{\top }$ representing the transposition of column vector $\boldsymbol{x}_{i,j}$ (Equation 2.).

\begin{align}
{z}_{i,j}\sim \mathrm{Bernoulli}\left({\psi}_{i,j}\right)\\
\mathrm{logit}\left({\psi}_{i,j}\right)={\boldsymbol{x}}_{i,j}^{\top}\boldsymbol{\beta_{i}}
\end{align}

The regression coefficients in these multi-species occupancy models are described as random effects arising from a common community level distribution (Equation 3.). Where $\boldsymbol{\mu}_{\beta}$ represents the community level mean effect for each occurrence covariate effect and $\boldsymbol{T}_{\beta}$ is a diagonal matrix representing the variability of these among species in the community.

\begin{align}
\boldsymbol{\beta}_i \sim \text{Normal}(\boldsymbol{\mu}_{\beta}, \boldsymbol{T}_{\beta})
\end{align}

The detection component estimates the unobserved ${z}_{i,j}$. Here, ${y}_{i,j,k}$ is the observed detection or non-detection of a species $i$, at site $j$, during replicate $k$ (Equation 4.). This is approached as arising from a Bernoulli process conditional on the true latent occurrence process ${p}_{i,j,k}$. The probability of a species being detected at a site, during a replicate (given it is present at site $j$), is a function of site and replicate specific covariates $V$ and a set of species-specific regression coefficients ${\alpha}_{i}$ (Equation 5.). 

\begin{align}
{y}_{i,j,k}\sim \mathrm{Bernoulli}\left({p}_{i,j,k}{z}_{i,j}\right)\\
\mathrm{logit}\left({p}_{i,j,k}\right)={\boldsymbol{v}}_{i,j,k}^{\top}\boldsymbol{\alpha}
\end{align}

Similarly to Equation 3. these coefficients are described as random effects arising from a common community level distribution, where $\boldsymbol{\mu}_{\alpha}$ represents the community level mean effect for each detection covariate effect and $\boldsymbol{T}_{\alpha}$ is a diagonal matrix representing the variability of these among species in the community (Equation 6.).

\begin{align}
\boldsymbol{\alpha}_i \sim \text{Normal}(\boldsymbol{\mu}_{\alpha}, \boldsymbol{T}_{\alpha})
\end{align}

Minimally informative priors were specified for community and species level coefficients (${\alpha}$ and ${\beta}$, a normal prior of mean = 0, variance = 2.72) and for community level occurrence and detection variance parameters ($\boldsymbol{T}_{\alpha}$ and $\boldsymbol{T}_{\beta}$, 0.1 for the scale and shape parameters of the inverse Gamma prior).

We included covariates in the model based on a pre-specified Directed Acyclic Graph and after assessing for co-linearity (defined as strong correlation >0.8) among variables (Supplementary Figure 2.). Continuous variables were standardised by scaling to values between 0-1. The fully specified model is defined in Equation 6 and 7.

$$
\text{Probability of occurrence} \sim \text{Landuse} + \text{Village} + \text{scale}(\text{Distance to permanent structure}) +\\ \text{scale}(\text{Distance to village centre}) + \text{scale}(\text{Elevation}) \tag{7}
$$
$$
\text{Probability of detection} \sim \text{scale}(\text{Monthly precipitation}) + \text{Moon fraction} + \text{scale}(\text{Number of trap nights}) \tag{8}
$$

Using this model, we calculate the effect of land-use as the difference in occupancy probability for each species between each of the four land use classifications. Only estimates for species with at least X records to avoid inference from limited data. Occupancy is interpreted here as the species' probability of being detected through a successful trapping event during the study. We estimate species richness in each habitat type by obtaining the sum of species at a trapping site for each iteration of the Bayesian sampling process to compare rodent assemblage responses to land use classification.

### Co-occurrence of species

To investigate the presence of competitive exclusion by rodent species in this setting we adopted a Binomial joint likelihood model fit using Integrated Nested Laplace Approximation (INLA) in the R statistical computing language [@rue_approximate_2009; @martins_bayesian_2013].Here we set up co-occurrence models to estimate the probability of detection of two species $Y_{a}$ and $Y_{b}$ (Equation 8). Here, ${\gamma}$ is the species specific intercept for species ${a}$ and ${b}$ respectively. ${A.co.B}$ and ${B.co.A}$ represent the co-occurrence of species ${a}$ with species ${b}$ and species ${b}$ with species ${a}$ respectively.

\begin{align}
{Y}_{a} \sim {\gamma}_{a} + \mathrm{A.co.B} \tag{9} \\ 
{Y}_{b} \sim {\gamma}_{b} + \mathrm{B.co.A} \tag{10}
\end{align}

Minimally informative priors were set for the intercept and species co-occurrence variables with a mean of 0 and precision of 1.

# Results

## Rodent occurrence and species assemblage structure

During the study period 749 small mammals were obtained from 40,152 trap-nights across the four study villages (1.9% trap-success (TS)). The greatest number of small mammals were detected in agricultural settings, these communities also had the greatest species richness and Shannon diversity. However, the greatest TS was obtained from traps set within households, although in these settings species richness and diversity was lower (Table 1a). Seilama and Lalehun had the highest overall TS, species richness and Shannon diversity although the land-use types with the highest values of these metrics differed between the villages. Lambayama the most urbanised village had the lowest species richness and Shannon diversity with the majority of rodents trapped within households.

\captionsetup[table]{labelformat=empty}

```{r table-1a, layout = "l-page", tab.cap = "Table 1a. The number of individual small mammals (N), the number of trap nights (TN), trap-success (TS %), species richness and Shannon diversity are presented for each village and landuse type."}

table_1a <- readRDS(here("output", "table_1a.rds"))
table_1a %>%
  as_grouped_data(groups = c("village")) %>%
  flextable() %>%
  autofit() %>%
  set_header_labels(landuse = "Landuse",
                    village = "Village",
                    N = "N",
                    TN = "TN (TS %)",
                    species_richness = "Species richness",
                    shannon_diversity = "Shannon diversity") %>%
  fit_to_width(6)

```

The most commonly detected rodent species was *M. natalensis* (N = 144, 19.2%), followed by *R. rattus* (N = 90, 12%), *M. musculus* (N = 86, 11.5%) and *Lophuromys rostratus* (N = 74, 9.9%) (Table 1b.). *M. natalensis* and *R. rattus* were detected in all villages with *M. natalensis* never detected in fallow or forest settings. Conversely, *Hybomys planifrons* and *Gerbilliscus kempii* were only detected in a single village and in forest and fallow settings respectively. The invasive rodent *M. musculus* was only detected in Lambayama and Seilama village settings. We found little variation in species richness by season, however, prevalence of species (measured by trap-success) fell for most species, except *M. natalensis* which was more commonly trapped (greater trap-success) during the wet season in indoor settings across the three village sites in which it was prevalent (Supplementary Table 1. **Not done yet**).

```{r table-1a, layout = "l-page", tab.cap = "Table 1b. The location of detection of individual small mammals by species. Values in paranthesis are the TS % for that species in the village and landuse for the associated traps."}

table_1b <- readRDS(here("output", "table_1b.rds"))
table_1b %>%
  as_grouped_data(groups = "clean_names") %>%
  flextable() %>%
  autofit() %>%
  set_header_labels(clean_names = "Species",
                    village = "Village") %>%
  fit_to_width(6)

```


## Rodent species networks

Rodent species networks varied by village and habitat type. The largest networks (greatest number of co-located species) were obtained from the villages Seilama and Lalehun.

## Estimating the effect of land use on species occurrence and richness

We identified landuse level and village level variability in the occurrence of small mammal species in our model incorporating incomplete detection. 

-   Model seasonal change in a) relative abundance of all rodents and b) number of species by seasonal indicators and proportions of different land use types as covariates
-   Species distribution maps will be produced
-   Difference between observed and expected will be investigated by comparing to observations of other species within these distributions

## Species distribution maps for current land use classifications and potential future change

-   Apply these species distribution models to expected future scenarios

# Discussion

# Conclusion

# References

# Supplementary figure 1.

```{r species-accumulation, layout = "l-page", fig.cap = "Supplementary figure 1. Species accumulation curves for each village site"}

knitr::include_graphics(here("output", "species_accumulation.pdf"))

```

```{r model-DAG, layout = "l-page", fig.cap = "Supplementary figure 2. Directed Acyclic Graph used for occupancy and detection model specification"}

knitr::include_graphics(here("output", "model_dag.pdf"))

```
